---
title: "Session 1"
subtitle: "Data Wrangling in R"
author: "Youth Impact R Workshop Series"
institute: "Zezhen Wu"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      countIncrementalSlides: false
      highlightLines: true
      highlightStyle: solarized-light
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(xaringanthemer)
library(magrittr)
library(flair)
library(tidyverse)
# xaringan::inf_mr()
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
# style_mono_accent(
#   text_font_google = google_font("Montserrat", "300", "300i"),
#   code_font_google = google_font("Fira Mono")
# )
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono")
)
```

<style>
:root {
  --font12: 12pt;
  --font10: 10pt;
}

.font12 > table {
  font-size: var(--font12);
}

.font12 > p {
  font-size: var(--font12);
}

.font12 > pre .remark-code-line {
  font-size: var(--font12);
}

.font10 > table {
  font-size: var(--font10);
}

.font10 > p {
  font-size: var(--font10);
}

.font10 > pre .remark-code-line {
  font-size: var(--font10);
}
</style>

# Disclaimer

Part of this presentation is borrowed from the public [GitHub repository](https://github.com/nyuglobalties/workshops) for the NYU Global TIES R workshop coauthored by Patrick Anker and Zezhen Wu.

---

# Agenda

By the end of this workshop, you should be able to understand the following code:

```{r, eval = FALSE}
library(tidyverse)

dat <- read_csv(here::here("data/RCT.csv"))

dat %>%
  filter(age <= 10) %>%
  group_by(sch_id) %>% 
  summarize(avg_scores = mean(std_scores)) %>% 
  mutate(transformed_scores = as.numeric(scale(avg_scores))) %>% 
  select(-avg_scores)
```

We'll be covering

- Basic R syntax
- Data loading
- Data inspection
- Data manipulation

---

# Type of R files

### R Projects (.Rproj)

- Organize work in a structured manner with relative path.
- Provide a dedicated workspace for each project.

### R Script (.R)

- Plain text files containing R code.
- Executed directly or sourced within other scripts or documents.

### R Markdown (.Rmd)

- Combine R code, text, and output.
- Create dynamic documents.

--

**We will talk about data workflow management in a future workshop.**

**Let's try to create these files by hand!**

---
class: middle, center

# Loading your data

## 🗄➡ 💻

---

## Understanding R syntax

R is built around **functions**, bits of atomic code that accept inputs and return an output.

--

A function looks like this:

```{r, eval=FALSE}
function_name(argument_1, argument_2, parameter = TRUE)
```

--

Functions can be bundled together and shared using **packages**. To access a function from a specific package, we use the `::` operator.

```{r, eval=FALSE}
package.name::function_name(argument_1, argument_2, parameter = TRUE)
```

---

## Understanding R syntax

--

It may be cumbersome to write `::` each time you use a function. Therefore, R allows you to load those packages in your current session using the `library()` function so you can just refer to the function name without `::`.

--

```{r, eval=FALSE}
library(package.name)

# Refers to `function_name` in `package.name`
function_name(argument_1, argument_2, parameter = TRUE)
```

--

In Stata, this would be similar to:

```
// function_name argument_1 argument_2, parameter 1
```

---

# Tools for loading data

While there are lots of options, I recommend loading data using the following functions:

```{r, eval=FALSE}
readr::read_csv()
haven::read_dta()
openxlsx::read.xlsx()
```

**readr** and **haven** are both part of the "_tidyverse_", a bundle of packages that contain the most essential functions for data wrangling.

**openxlsx** is a standalone package that deals with reading and writing Excel files.

---


# Tools for loading data

--

Some packages come with the installation of R, some don't. This is because R is an open-source software, and anyone in the world can contribute their packages to the R community.

--

To use these packages, you need to install them first by running `install.packages("package.name")`:

```r
install.packages("tidyverse")
install.packages("haven")
```

**Note:** Only run these commands **once**. Once they're installed, you don't need to run that command again since the package will be on your computer.

---

# Finding help on packages

R packages are typically published on [CRAN](https://cran.r-project.org/web/packages/available_packages_by_name.html) (The Comprehensive R Archive Network).

Developer versions of R packages are typically published on GitHub (for instance, [`readr`](https://github.com/tidyverse/readr)).

When you encounter a function that you don't know or understand for some reason, **consult the R help system**!

--

```{r, eval=FALSE}
# search help doc for a package
?readr

# search help doc for a function in a package
?readr::read_csv

# If you're not certain what a function name is or you're looking for a certain topic, use `??` (think: "wildcard search") followed by the topic name
??readr::read_table
```

---


# Let's load our data!

The simulated RCT dataset is located in the `data` directory (folder) at the root of this project. I'll use the **here** package to create paths that are _relative to the project root_.

--

I'll attach the packages we need for now:

```{r}
library(readr)
library(here)
```

Note how **here** tells us where it thinks the project root is! It recursively searches upward for a directory with a `DESCRIPTION` or `.Rproj` file. 

**here** is very handy utility for reproducibility since it _localizes_ paths to the project -- and not expanding to the full path on _your_ computer.

---

# Let's load our data!

Time to load the data:

```{r}
dat <- read_csv(here("data/RCT.csv"))
```

---

## Two things to note

.pull-left[

R uses the `<-` operator to _assign_ values to symbols (variables). 

```{r, eval=FALSE}
x <- (1+1) * 2^3 / 4 - 2
```

In our data's case, the results of reading the CSV file are assigned to the `dat` symbol.

]

.pull-right[

Unlike Stata, you can have multiple data frames in one R session.

```{r, eval=FALSE}
dat1 <- dat
dat2 <- dat
```

This is very helpful in cross-checking and merging datasets, as well as interacting with various R objects (e.g., vectors, matrices, lists, data frames, etc.).

]

---

# Looking at our data

After loading our data, it's useful to high-level view of its content to make sure that it loaded correctly.

We'll use the `str()` and `print()` functions here, both part of base R.

---

# Looking at our data

```{r, include=FALSE}
attr(dat, "spec") <- NULL
```

```{r}
str(dat)
class(dat)
```

---

# Looking at our data

The _class_ of this `dat` object is a `data.frame`. `data.frame`s are R's representation of tabular data. `dat` is also a `tbl_df` ("tibble"), which is the Tidyverse's version of a `data.frame` that is a little bit more stringent (i.e. safe) than base R's version, and it has a prettier `print()` method:

```{r}
print(dat)
```

---

# Looking at our data

If you'd like to see _all_ of your data in a pretty Excel-like spreadsheet, use the `View()` function which has a great display in RStudio.

```{r, eval=FALSE}
View(dat)
```

Note: Or you can click on the data frame in the Environment tab. Or hold command/ctrl and click `dat` on the R script.

---

# Data types in R

You may have noticed a bunch of three-letter abbreviations next to the variable names in `print(dat)` and the variable content in `str(dat)`. These are the _type signatures_ for each variable:

- Double (`dbl` / `num`): All real numbers, like $50$, $23.193$, or $1.24\times10^{-8}$
- Character (`chr`): Text/strings, written with single- or double-quotes (e.g. `"a string"`)

There are a couple more that you will encounter regularly:

- Integer (`int`): Integers and counting numbers. No decimals and scientific notation here!
- Logical (`lgl`): Boolean data, written as `TRUE` or `FALSE`.
- Factor (`fct`): Factors, R's representation of categorical data. They look like character data with `print()`, but under the hood they are integers. These should only exist in cleaned, ready-for-analysis data!

---

# Quiz

.pull-left[

How do I create a variable `x` that refers to the sentence "Hello, world!" ?

]

.pull-right[

Use 1, 2, 3, or 4 to refer to the code line number.

```{r, eval=FALSE}
x "Hello, world!"
"x" <- Hello, world!
x <- "Hello, world!"
x <- Hello, world!
```

]

---

# Quiz

.pull-left[

How do I create a variable `x` that refers to the sentence "Hello, world!" ?

]

.pull-right[

Use 1, 2, 3, or 4 to refer to the code line number.

```{r, eval=FALSE}
x "Hello, world!"
"x" <- Hello, world!
*x <- "Hello, world!"
x <- Hello, world!
```

]

---
class: middle, center

# Working with your data

## 🗄🔍

---

# Working with your data

What we'll be discussing today is basic *data wrangling*.

> Data wrangling (n.): The process of transforming data from one raw form into another format with the intent of making it more appropriate and valuable for a variety of downstream purposes such as analytics.
> 
> <em>&mdash;Wikipedia</em>

---

# Basic Data Wrangling

Most data wrangling comes down to **five** basic operations:

- **Selecting**: Selecting or deselcting variables 
- **Filtering**: Filtering by row
- **Arranging**: Reordering by variables
- **Summarizing**: Aggregating data for summary statistics
- **Mutating**: Creating, editing, or deleting variables

You can also **Group** these operations by variables. For example, finding the average math score in each class would be achieved by _grouping_ by class and then _summarizing_ by taking the mean.

---

# Basic Data Wrangling

In this workshop, **we recommend using the dplyr package in tidyverse**, but we can also achieve the same thing in base R:

.font12[

| Operation   |                                       Tidyverse |                                    Base R       |
|:------------|------------------------------------------------:|------------------------------------------------:|
| Selecting   | `select(dat, treated)`                          | `dat[,"treated"]`                               |
| Filtering   | `filter(dat, treated == 1)`                     | `dat[dat$treated == 1", ]`                      |
| Arranging   | `arrange(dat, std_scores)`                      | `dat[order(dat$std_scores), ]`                  |
| Summarizing | `summarize(dat, avg_scores = mean(std_scores))` | `data.frame(avg_scores = mean(dat$std_scores))` |
| Mutating    | `mutate(dat, failed = std_scores < 0)`          | `dat$failed <- dat$std_scores < 0`              |

]

It's  apparent that tidyverse has a much more consistent and readable syntax than base R.

---

# Basic Data Wrangling

All of these tidyverse functions are provided by the **dplyr** package.

```{r}
library(dplyr)
```

---

## Selecting

Selects or deselects one or multiple variables.

```{r, eval = F}
select(dat, treated)
```

```{r, echo = F}
head(select(dat, treated), 1)
```

```{r, eval = F}
select(dat, -treated)
```

```{r, echo = F}
head(select(dat, -treated), 1)
```

```{r, eval = F}
select(dat, sch_id:female)
```

```{r, echo = F}
head(select(dat, sch_id:female), 1)
```
---

## Filtering

Looks for rows with a provided comparison, usually with `<`, `>`, `<=`, `>=`, or `==` (note: two equals!).

```{r}
filter(dat, treated == 1)
```

**Homework: Figure out how to choose row-wise those with age greater than the average age.**

---

## Arranging

Sorts a dataset by a `variable` in _ascending_ order with `arrange(dat, variable)`.

```{r, eval=FALSE}
arrange(dat, std_scores)
```

```{r, echo=FALSE}
head(arrange(dat, std_scores), 6)
```

**Homework: Figure out how to order by _descending_ order.**

---

## Summarizing

Calculate some summary function on variables, like `mean()`, `median()`, or `sd()`:

```{r}
summarize(
  dat, 
  avg_score = mean(std_scores), 
  sd_score = sd(std_scores)
)
```

**Homework: Is there a difference between summarize() and summarise()?**

---

## Mutating

Create, edit, or delete variables in a dataset.

```{r, eval=FALSE}
mutate(
  dat,
  # Changes condition to an index variable with character values
  condition = ifelse(treated == 1, "Treatment", "Control")
)
```

```{r, echo=FALSE}
head(mutate(
  dat,
  # Changes condition to an index variable with character values
  condition = ifelse(treated == 1, "Treatment", "Control")
), 2)
```

**Homework: Create a variable in `dat` called age_gt_10, which takes the value of 1 when age > 10, otherwise 0.**

---

.center[
# ⚠️
## Tidyverse Caution!
]

The tidyverse (mainly **dplyr**, but a couple other packages) refers to variables in datasets with "naked" symbols.<sup>1</sup>

```{r flaired_symbols, include=FALSE}
summarize(dat, avg_score = mean(std_scores))
```

```{r, echo=FALSE}
decorate("flaired_symbols") %>% 
  flair("std_scores")
```

.footnote[
[1]: In computer science lingo, these are technically called _quoted symbols_, but that can be confusing.
]

---

.center[
# ⚠️
## Tidyverse Caution!
]

But _outside_ of the tidyverse functions, these variables **are not defined**.

```{r, error=TRUE}
mean(std_scores)
```

---

.center[
# ⚠️
## Tidyverse Caution!
]

To access dataset variables outside of tidyverse functions, you must use the **`$`** ("extract") operator in conjunction with the symbol for the parent dataset.

```{r extract_op, include=FALSE}
mean(dat$std_scores)
```

```{r, echo=FALSE}
decorate("extract_op") %>% 
  flair_rx("\\w+\\$")
```

Just keep this in mind!

---

# Combining Operations

These "verbs" (as **dplyr** likes to call them) are **composable** meaning you can chain these commands together, allowing you to create complex data transformations or summaries all in one step.

```{r}
summarize(
  filter(dat, treated == 1), 
  avg_score = mean(std_scores)
)
```

--

This is ugly. Thankfully, there is a better way of chaining these steps together!

---

.center[

.font12[Introducing...]

]

--

.center[

## The Pipe!

# `%>%`

]

--

The pipe `%>%` is an operator provided by **magrittr**<sup>1</sup> that makes writing nested functions _much_ easier.

The shortcut is `command + shift + M` on Mac, or `ctrl + shift + M` on Windows.

.footnote[
[1]: Starting from R 4.1.0, there is the native pipe |> in R, which does the same thing without loading `tidyverse`.
]

---

## The Pipe

Instead of writing

```r
h(g(f(x)))
```

you can write

```r
x %>% 
  f() %>% 
  g() %>% 
  h()
```

--

The pipe works by inserting the result of the left hand side into the **first argument** of the right hand side.

---

# Combining Operations

Revisiting our first chained example,

```{r, eval=FALSE}
summarize(
  filter(dat, treated == 1), 
  avg_scores = mean(std_scores)
)
```

we can now use

```{r}
dat %>% 
  filter(treated == 1) %>% 
  summarize(avg_scores = mean(std_scores))
```

---

## Grouping

Now let's add grouping into the mix, which you can easily do with `group_by(variable)`:

```{r}
dat %>% 
  group_by(treated) %>% 
  summarize(avg_scores = mean(std_scores))
```

---

## Grouping

You can group by multiple variables as well. Just add more variables to `group_by()`:

```{r, eval=FALSE}
dat %>% 
* group_by(treated, female) %>% 
  summarize(avg_scores = mean(std_scores))
```

```{r, echo=FALSE}
dat %>% 
  group_by(treated, female) %>% 
  summarize(avg_scores = mean(std_scores))
```

---

# Recap

.pull-left[

.font12[

How would I find the median scores by treatment and gender?
]

]

.pull-right[

.font10[

```{r, eval=FALSE}
# 1.
dat %>% 
  filter(treated, female) %>% 
  summarize(
    med_scores = median(std_scores)
  )

# 2.
dat %>%
  summarize(
    med_scores = median(std_scores)
  ) %>% 
  group_by(treated, female) 

# 3.
dat %>%
  group_by(treated, female) %>% 
  summarize(
    med_scores = median(std_scores)
  )

# 4.
dat %>%
  summarize(
    med_scores = median(std_scores)
  ) %>% 
  filter(treated, female)
```

]

]

---

# Recap

.pull-left[

.font12[

How would I find the median scores by treatment and gender?

]

]

.pull-right[

.font10[

```{r, eval=FALSE}
# 1.
dat %>% 
  filter(treated, female) %>% 
  summarize(
    med_scores = median(std_scores)
  )

# 2.
dat %>%
  summarize(
    med_scores = median(std_scores)
  ) %>% 
  group_by(treated, female) 

# 3.
*dat %>%
* group_by(treated, female) %>% 
* summarize(
*   med_scores = median(std_scores)
* )

# 4.
dat %>%
  summarize(
    med_scores = median(std_scores)
  ) %>% 
  filter(treated, female)
```

]

]

---

## Bonus Point

**You can do Tableau-style plots in R using just one command:**

```{r, eval = F}
install.packages("esquisse")
library(esquisse)

# Open Tableau-like dashboard in Rstudio
esquisser()
```

**Note: We can talk about summary statistics and visualization in a future workshop.**

---

# Exercise

Use the functions you learned to answer the following question:

1. What is the average age for boys and girls?
2. Which `st_id` is the oldest students in school 8?
3. Which age group has the most number of students?
4. What is the average `std_scores` for girls aged 7 or 8 in the treatment group?
5. Create a data.frame with the name `dat_s` with the following rules:
- Deselect `st_id`
- Filter out students younger than 10 years old
- Create a new variable called `transformed_scores` and assign it the value of 99 times the `std_scores` if age >= 10, else assign it the value 101 times the `std_scores`

.footnote[

Consider using the `|` (read: **or**) operator to combine multiple filter statements into one (e.g. filtering for `this` or `that` would be `filter(dat, this | that)`).

]


